---
title: "Solutions to Exercise Sheet 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(VIM)
library(nycflights13)
```
##########################
Q1
Find all flights that:  
(a)	Had an arrival delay of two or more hours  
(b)	Flew to Houston (IAH or HOU)  
(c) Departed in summer (July, August, and September)  
(d)	Were delayed by at least an hour, but made up over 30 minutes in flight  
(e) Departed between midnight and 6am (inclusive)  

```{r, echo = TRUE, warning = FALSE}
Q1a <-filter(flights, dep_delay >= 2)

Q1b <-filter(flights, dest=="IAH" | dest== "HOU")

Q1c <- filter(flights, month == 7 | month == 8 | month ==9)
Q1c <- filter(flights, between(month, 7, 9))

Q1d <-mutate(flights, made_up = dep_delay - arr_delay) %>%
        filter(dep_delay >= 60 & made_up >30)

Q1e <-filter(flights, between(dep_time, 0, 600) | dep_time == 2400)

```


Q2
How many flights have a missing dep_time? What other variables are missing? 

We find that the data set containing observations where dep_time is missing has length 8255. We see that observations missing the dep_time are also missing arr_time, dep_delay, arr_delay and air_time.

```{r, echo = TRUE }
Q2 <-filter(flights, is.na(dep_time))
```

Q3
Convert dep_time and sched_dep_time to represent the number of minutes since midnight.

The floor function will round down to the nearest integer.
```{r, echo = TRUE}
Q3<-mutate(flights,
       dep_time = 60 * floor(dep_time/100) + (dep_time - floor(dep_time/100) * 100),
       sched_dep_time = 60 * floor(sched_dep_time/100) + (sched_dep_time - floor(sched_dep_time/100) * 100))

```
Q4

Find the 10 most delayed flights using a ranking function. 
The most delayed flights have the largest values for dep_delay. The rank() function ranks from smallest to largest (smallest number gets rank 1).
The desc() function tranforms a vector into a format that will be sorted in descending order. We can use this with the filter command to select the observations where dep_delay has a rank of 10 or less.

```{r, echo = TRUE,}
Q4<-filter(flights, rank(desc(dep_delay))<=10)

```
Q5	
Download the Pb_messy.csv and Pb_tidy.csv data sets from Blackboard.
Can you create an R script that will transform the messy data set into the tidy data set?

First read in the data and send it to the gather function using the pipe.
Next create two new columns "site" and "Pb" and gather the columns.
The key is the name of the variable that forms the column names i.e. site 
and the values are spread over the cells i.e. levels of Pb.
Use separate() to split the site column into two new columns, one showing site (1 or 2), the other showing the plot (A or B).

```{r, echo = TRUE, message=FALSE}
  Pb_tidy <-read_csv("Pb_messy.csv") %>% 
  gather("site", "Pb", 2:5) %>%         
  separate("site", c("site", "plot"), sep = "_" )

write_csv(Pb_tidy,"Pb_tidy.csv")
```


Q6

Age has 4 missing values and Income has 5 missing values.	There is a missing data pattern for the Income variable, data is more likely to  be missing from Income when Age is high. There is no missing data pattern from Age.				
  
Q7

Read in the cholesterol data set and summarise
```{r, echo = TRUE, message = FALSE}
cholesterol<-read_csv("cholesterol.csv")
summary(cholesterol)
```


There are 7 variables. Age, Cholesterol, SysolicBP and BMI are continuous variables. Gender, Smoking and Education are categorical (the variable type is character). Notice that there are 16 missing data in the Cholesterol variable.

If our data set is stored as a Tibble, then categorical variables are simply columns containing characters. If we wish to use the catagorical variables in an alalysis then we need to convert them to a type of object called a factor. 
To convert the categorical variables (currently stored as characters) to factors
```{r, echo =TRUE}
cholesterol$Gender <- as_factor(cholesterol$Gender)
cholesterol$Smoking <- as_factor(cholesterol$Smoking)
cholesterol$Smoking <- as_factor(cholesterol$Smoking)
```


Next we plot histograms for all 4 variables.


```{r Histograms, echo = TRUE, fig.height=5, fig.width=4}
for(i in c(1,3,4,5)) hist(cholesterol[[i]], xlab = names(cholesterol[i]), main = names(cholesterol[i]))
```

The distribution of all the variables apart from SystolicBPlook normal.  
SystolicBP looks skewed a transformation may help but the number of large
values seems small on the histogram so lets investigate further using boxplots.

```{r, echo = TRUE,fig.height=5, fig.width=4}
 
for(i in c(1,3,4,5)) boxplot(cholesterol[[i]]~cholesterol$Gender, xlab = "Gender", main = names(cholesterol[i]))

```

The box plots above show that distribution of the continuous variables are similar for males and females, etc.
There appear to be some potential outliers, one particular value for SystolicBP looks very high - it has a value of 1750 - this is not a valid measurement for blood pressure so it could be replaced with an NA.
We can replace the value of 1750 with an NA and recheck the distributions 

```{r, echo = TRUE,fig.height=5, fig.width=4}
cholesterol<- cholesterol %>% mutate(SystolicBP = (replace(SystolicBP, SystolicBP==1750, NA)))
boxplot(cholesterol$SystolicBP~cholesterol$Gender, xlab = "Gender", main = "Systolic Blood Pressure")
```
Note that the skew has gone from SystolicBP

The scatterplots below don't show any relationships between the continuous variables.
```{r, echo = TRUE}
pairs(cholesterol[,c(1,3,4,5)])
```

The aggr function gives us an overview of the missing data.

```{r, echo = TRUE}
a <- aggr(cholesterol)
```


There are 16 missing data in Cholesterol and 1 in SystolicBP
To check for missing data patterns we can use the histMiss() and marginplot() functions for the continuous variables and the barMiss() function for the categorical variables. 
First need to convert the cholesterol tibble to a dataframe to use some of the functions in the VIM package.
```{r, echo = TRUE, message=FALSE}
cholesterol<-as.data.frame(cholesterol)

barMiss(cholesterol[,c("Gender", "Cholesterol")])
barMiss(cholesterol[,c("Education", "Cholesterol")])
barMiss(cholesterol[c("Smoking","Cholesterol")])
```

There appear to be more missing cholesterol values for female patients
```{r, echo = TRUE}

by(cholesterol$Cholesterol,cholesterol$Gender, summary)

```

Out of a total of 250 observations (143 female and 107 male) there are 10 missing values for females and 6 for males. We can test to see whether there is a significant difference using prop.test. If you have not used this test before, be sure to look up the help for this function to find out  how it works.
```{r, echo = TRUE}
prop.test(c(10,6), c(143,107))
```
We conclude that difference is not unusually large. The data does not support the hypothesis that Gender impacts the amount of missing data in the Cholesterol variable.

Next we check for missing data patterns between Cholesterol and the other continuous variables.

```{r, echo = TRUE}
histMiss(select(cholesterol, "BMI","Cholesterol"))
histMiss(select(cholesterol, "SystolicBP","Cholesterol"))
histMiss(select(cholesterol, "Age","Cholesterol"))
```

There appears to be a missing data pattern for Cholesterol and BMI
where data is more likely to be missing from Cholesterol when BMI is high. 
 
Q8

(a) The data missing from the census questionnaire is MAR since it 
is related to whether there are children in the household, if there are no children in the house hold then the data is missing. This missing data process is ignorable and no action is taken.

(b) The data missing for the electoral poll questionnaire are MCAR since the participants are chosen at random. Provided the sample size is large enough, the data can be analysed using complete cases only.

(c) The data missing for the concentration of pesticide is MNAR since the 
data is missing for the plants that received the highest concentration of
pesticide. In this instance knowledge of the toxisity of the pesticide at 
high concentrations should be presented in the analysis.

